Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-19 doc/mc-paper.tex
adjust phrasing in introduction

\emph{Knowledge compilation} is the process of converting a Boolean
formula into a form that makes it possible to calculate associated
values efficiently, such as the weighted or unweighted model
count. Existing knowledge compilers
provide no guarantee that the representation they produce
correctly represents the original formula, and therefore they cannot validate the generated counts.

@@

We have developed a program that generates POG representations from
the d-DNNF graphs produced by the state-of-the-art knowledge compiler
\dfour, as well as checkable proofs that the output POGs
are equivalent to the input CNF formulas.  We have also implemented an
efficient checker for these proofs. Our tool chain
for generating and verifying POGs scales to all but the largest d-DNNF
graphs produced by \dfour{} for formulas from a recent model counting
competition. We also present a checker and a model counter for POGs that have
been formally verified in the \lean{} proof assistant.  These programs both prove the soundness of our proof framework and
create the first formally verified toolchain for weighted and unweighted model counting.
\end{abstract}

@@

Modern Boolean satisfiability (SAT) solvers can
find a satisfying assignment to a Boolean formula or generate a proof that no
such assignment exists.  They have applications across a variety of
domains, including computational mathematics, combinatorial
optimization, and the formal verification of hardware, software, and
security protocols.  Some applications, however, require going
beyond standard Boolean satisfiability.  For example, the {\em model
  counting problem} requires computing the number of satisfying
assignments of a formula, including in cases where there are far too many
to enumerate individually.  Model counting, in turn, has
applications in artificial intelligence, computer security, and
statistical sampling.  There are also many useful extensions of standard model counting,
including {\em
  weighted model counting}, where a weight is defined for
each possible assignment, and the goal becomes to compute the sum of the weights
of the satisfying assignments.

Model counting is a challenging problem---more challenging that the
already NP-hard problem of Boolean satisfiability.  Several
tractable variants of Boolean satisfiability, including 2-SAT, become
intractable when the goal becomes model counting and not just
satisfiability \cite{valiant:siam:1979}.  Nonetheless, a number of
model counters have developed that scale to very large formulas, as
witnessed by the progress in recent model counting competitions.

One approach to model counting, known as {\em knowledge compilation},
transforms the formula into a structured form for which model counting
becomes straightforward.  For example, the {\em deterministic,
  decomposable, negation normal form} (d-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by Boolean operation And or Or.  Restrictions
are placed on the structure of the graph such that a count of the
models can be computed by a single bottom-up traversal.
Kimmig et al.\ present a very general framework, known as {\em
  algebraic model counting}~\cite{kimmig:jal:2017}, describing
properties of Boolean functions that can be efficiently computing from
a d-DNNF representation.  These include standard and weighted model
counting, and much more.

One shortcoming of existing knowledge compilers is that they have no
way to validate that their generated results are correct, i.e., that
the compiled representation is logically equivalent to the original
formula.  By contrast, all modern SAT solvers can generate
checkable proofs when they encounter unsatisfiable formulas.  The
guarantee provided by a checkable certificate of correctness enables
users of SAT solvers to fully trust their results.  Experience has also
shown that having this capability gives SAT solver developers an ability to quickly
detect and diagnose bugs in their programs, and this, in turn, has led
to more reliable SAT solvers.

In this paper, we present \emph{Partitioned Operation Graphs} (POGs),
a generalization of the representations produce by current knowledge
compilers, including d-DNNF\@.  We present the CRAT file format.
It captures both the structure of a POG and a checkable proof
that the POG is logically equivalent to a Boolean formula given in
conjunctive normal form (CNF).  A proof consists of a sequence of
clause addition and deletion steps, based on an extended resolution
proof system~\cite{Tseitin:1983}.  We present a set of conditions that
a CRAT file must satisfy to provide an equivalence proof.

@@

\end{tikzpicture}
}
\caption{Certifying Tool Chain.
  The output of a standard knowledge compiler is converted into a combined graph/proof (CRAT) that can be independently checked and evaluated}
\label{fig:chain}
\end{figure}

Figure~\ref{fig:chain} illustrates the certifying knowledge compilation
and (weighted) model counting tool chain that we have developed.  A
{\em proof generator} generates a CRAT description from a d-DNNF
graph produced by \dfour{}, a state-of-the-art knowledge compiler.
A {\em proof checker} then verifies that the combination of CNF and CRAT files
satisfy our rules, and a {\em ring evaluator} can compute
either a standard or weighted model count from the POG\@ representation.
We evaluate the capabilities of this tool chain using
benchmark formulas from the 2022 standard and weighted model
competitions.  Our tools are able to handle all but the largest d-DNNF
graphs generated by \dfour{}.
As the dashed box in Figure~\ref{fig:chain} indicates, this tool chain
shifts the need for trustworthy code away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will also be caught by the proof checker.

To further enhance the trustworthiness of our toolchain, we have
developed formally verified versions of the proof checker and ring
evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.  Running these two programs
has the effect of formally verifying both the logical equivalence of
the POG to the input formula and the correctness of the count.  In
addition, the existence of a formally verified checker serves the
important role of establishing the soundness of our proof framework.
Our
experience with developing a formally verified proof checker has shown
us that, even within the well-understood framework of extended
resolution, it can be challenging to formulate a full set of requirements to guarantee soundness.
Our efforts to formalize the framework
identified subtle details for some of the proof obligations.

@@

Having a SAT solver generate a proof of unsatisfiability has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently (and, in some cases, formally) checked
\cite{cruz-cade-2017,RAT,Tan:2021,wetzler14_drattrim}.  Some progress
has also been made on extending these proof systems to other domains,
including quantified Boolean
formulas~\cite{bryant:cade:2021,heule:JAR2014}.

@@

Our tool chain is not directly
comparable to theirs---they certify the output of a standard modeling
program, while we certify the output of a knowledge compiler and then
produce a certified standard or weighted model count.
%% Their tool
%% chain can be applied to some programs that are pure model counters,
%%while ours can be applied to pure knowledge compilers.
Nonetheless,
we compare the performance of our tool chain to theirs as part of our
experimental evaluation in Appendix~\ref{app:experiments}.

@@

%%Figure~\ref{fig:chain} illustrates one method of generating a CRAT
%%file from an input formula in conjunctive normal form.  Our tool chain
%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
%%but it would also work for other tools that use a top-down search
%%strategy to generate a d-DNNF representation.
Converting a d-DNNF
graph into a POG is straightforward---each node in a d-DNNF graph can
be directly represented as a POG node, although
our program performs simplifications to
eliminate Boolean constants.
Except in degenerate cases,
%where the formula is unsatisfiable or a tautology,
we can therefore assume
that the POG does not contain any constant nodes.
In addition, negation is only
applied to variables, and so we can merge the negation operations into the terminal nodes,
viewing the POG as consisting
of {\em literal} nodes corresponding to input variables and their negations, along with
{\em nonterminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.

>>>>

\emph{Knowledge compilation} is the process of converting a Boolean
formula into a form that makes it possible to efficiently calculate
associated values, such as the weighted or unweighted model
count. Existing knowledge compilers
provide no guarantee that the representation they produce
correctly represents the original formula, and therefore they cannot validate the generated counts.

@@

We have developed a program that generates POG representations from
the d-DNNF graphs produced by the state-of-the-art knowledge compiler
\dfour, as well as checkable proofs that the output POGs
are equivalent to the input CNF formulas.  We have also implemented an
efficient checker for these proofs. Our toolchain
for generating and verifying POGs scales to all but the largest d-DNNF
graphs produced by \dfour{} for formulas from a recent model counting
competition. We also present a checker and a model counter for POGs that have
been formally verified in the \lean{} proof assistant.  These programs both prove the soundness of our proof framework and
create the first formally verified toolchain for weighted and unweighted model counting.
\end{abstract}

@@

Given a Boolean formula $\phi$, modern Boolean satisfiability (SAT) solvers can
find an assignment satisfying $\phi$ or generate a proof that no
such assignment exists.  They have applications across a variety of
domains including computational mathematics, combinatorial
optimization, and the formal verification of hardware, software, and
security protocols.  Some applications, however, require going
beyond Boolean satisfiability.  For example, the {\em model
  counting problem} requires computing the number of satisfying
assignments of a formula, including in cases where there are far too many
to enumerate individually.  Model counting has
applications in artificial intelligence, computer security, and
statistical sampling.  There are also many useful extensions of standard model counting,
including {\em
  weighted model counting}, where a weight is defined for
each possible assignment, and the goal becomes to compute the sum of the weights
of the satisfying assignments.

Model counting is a challenging problem---more challenging than the
already NP-hard Boolean satisfiability.  Several
tractable variants of Boolean satisfiability, including 2-SAT, become
intractable when the goal is to count models and not just determine
satisfiability \cite{valiant:siam:1979}.  Nonetheless, a number of
model counters that scale to very large formulas have been developed, as
witnessed by the progress in recent model counting competitions.

One approach to model counting, known as {\em knowledge compilation},
transforms the formula into a structured form for which model counting
is straightforward.  For example, the {\em deterministic,
  decomposable, negation normal form} (d-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by a Boolean And/Or operation.  Restrictions
are placed on the structure of the graph such that a count of the
models can be computed by a single bottom-up traversal.
Kimmig et al.\ present a very general {\em
  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
properties of Boolean functions that can be efficiently computed from
a d-DNNF representation.  These include standard and weighted model
counting, and much more.

One shortcoming of existing knowledge compilers is that they have no
way to validate that their generated results are correct, i.e., that
the compiled representation is logically equivalent to the original
formula.  By contrast, all modern SAT solvers can generate
checkable proofs when they encounter unsatisfiable formulas.  The
guarantee provided by a checkable certificate of correctness enables
users of SAT solvers to fully trust their results.  Experience has also
shown that proof production capabilites allow SAT solver developers to quickly
detect and diagnose bugs in their programs. This, in turn, has led
to more reliable SAT solvers.

In this paper, we introduce \emph{Partitioned Operation Graphs} (POGs),
a generalization of the representations produced by current knowledge
compilers, including d-DNNF\@. To capture both the structure of a POG
and a checkable proof of its logical equivalence to a Boolean formula in
conjunctive normal form (CNF), we propose the CRAT file format. A CRAT
proof consists of a sequence of clause addition and deletion steps,
based on an extended resolution proof system~\cite{Tseitin:1983}.
We establish a set of conditions that a CRAT file must satisfy in order to
encode a well-formed POG and provide a valid equivalence proof.

@@

\end{tikzpicture}
}
\caption{Certifying toolchain.
  The output of a standard knowledge compiler is converted into a combined graph/proof (CRAT)
  which can be independently checked and evaluated}
\label{fig:chain}
\end{figure}

Figure~\ref{fig:chain} illustrates our certifying knowledge compilation
and (weighted) model counting toolchain.  A
{\em proof generator} generates a CRAT description from a d-DNNF
graph produced by \dfour{}, a state-of-the-art knowledge compiler.
A {\em proof checker} then verifies that the combination of CNF and CRAT files
satisfy our rules, and a {\em ring evaluator} can compute
either a standard or weighted model count from the POG\@ representation.
We evaluate the capabilities of this toolchain using
benchmark formulas from the 2022 standard and weighted model
competitions.  Our tools are able to handle all but the largest d-DNNF
graphs generated by \dfour{}.
As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
moves the root of trust away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will be caught by the proof checker.

To ensure soundness of the abstract CRAT proof system, as well as correctness of its
concrete implementation, we formally verified the proof system as well as versions
of the proof checker and ring evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.
Running these two programs on a particular CRAT file gives assurance that the proof
and the model count are correct. Our experience with developing a formally verified
proof checker has shown that, even within the well-understood framework of extended
resolution, it can be challenging to formulate a full set of requirements that guarantee soundness.
In fact, subtle details in the partitioned sum rule were identified during this process.

@@

Generating proofs of unsatisfiability in SAT solvers has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently (and, in some cases, formally) checked
\cite{cruz-cade-2017,RAT,Tan:2021,wetzler14_drattrim}.  Some progress
has also been made on extending these proof systems to other domains,
including quantified Boolean
formulas~\cite{bryant:cade:2021,heule:JAR2014}.

@@

Our toolchain is not directly
comparable to theirs---they directly certify the output of a model counting
program, while we certify the output of a knowledge compiler and then
produce a certified model count.
%% Their tool
%% chain can be applied to some programs that are pure model counters,
%%while ours can be applied to pure knowledge compilers.
Nonetheless,
we compare the performance of our toolchain to theirs as part of our
experimental evaluation in Appendix~\ref{app:experiments}.

@@

%%Figure~\ref{fig:chain} illustrates one method of generating a CRAT
%%file from an input formula in conjunctive normal form.  Our toolchain
%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
%%but it would also work for other tools that use a top-down search
%%strategy to generate a d-DNNF representation.
Converting a d-DNNF
graph into a POG is straightforward---each node in a d-DNNF graph can
be directly represented as a POG node, although
our program performs simplifications to
eliminate Boolean constants.
Except in degenerate cases,
%where the formula is unsatisfiable or a tautology,
we can therefore assume
that the POG does not contain any constant nodes.
In addition, negation is only
applied to variables, and so we can merge the negation operations into the terminal nodes,
viewing the POG as consisting
of {\em literal} nodes corresponding to input variables and their negations, along with
{\em nonterminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.
